Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2018-12-03 article_springer/article_springer.Rmd
add stability plots for Example 1 and Example 2

Since the exponential distribution is nested in the Mittag-Leffler family of distributions, an appropriate way to choose between a model with exponential and Mittag-Leffler inter-exceedance times seems to be a Likelihood ratio test. Although the two models are nested, the assymptotic distribution is not $\chi^2_1$-distributed, the classical Theorem of Wilk doesn't apply since under $H_0$ the paremter $\beta$ of Mittag-Leffler distribution is equal to zero and hence lies under $H_0$ on the boundary of the parameter space $(0,1]$. But one can perform a bootstrapped log-likelihood ratio test. In Figure \ref{Fig:LRT} one can see the power for the bootstrapped LRT for Mittag-Leffler distributions with different tail parameters. As expected the power gets smaller for tail parameters close to one, since the Mittag-Leffler distribution converges for $\beta \rightarrow 1$ to an exponential distribution. Hence, for tail parameters close to one it is hard to differentiate a Mittag-Lefller distribution from an exponential.

@@

The Theorem in Section \ref{sec:scaling} implies that for a high 
threshold $\ell$ we may 
approximate the distribution of $T(\ell)$ with an 
${\rm ML}(\beta, b(1/p))$ distribution, 
where $b(\,)$ varies regularly at $\infty$ with parameter $1/\beta$.
Building on the POT (Peaks Over Threshold) method, we propose the 
following estimation procedure for the distribution of $T(\ell)$:

@@

According to Figure \ref{fig:flares}, for a threshold $\ell$ at the $k$-th order 
statistic, the fitted threshold exceedance time distribution is 
$$
T_\ell \sim {\rm ML}(\beta, k^{1/\beta} \sigma_0), 
$$
where $\beta = 0.85$ and $\sigma_0 = 3.0 \times 10^7 {\rm sec}$. 
Unlike the exponential distribution, the Mittag-Leffler distribution is not memoryless, and the probability density of the time $t$ until the next threshold crossing will depend on the time $t_0$ elapsed since the last threshold crossing.
This density equals 
$$
p(t|\beta, \sigma_0, \ell, t_0) = \frac{f(t + t_0 | \beta, k^{1/\beta} \sigma_0)}{\mathbf P[T_\ell > t_0]}
$$
where $f(\,\cdot\, | \beta, k^{1/\beta} \sigma_0)$ is the probability density of ${\rm ML}(\beta, k^{1/\beta} \sigma_0)$. 
The more time has passed without a threshold crossing, the more the probability distribution shifts towards larger values for the next crossing
(see Figure \ref{fig:hazard}, left panel). The hazard rate 
$$
h(t) = \frac{f(t| \beta, k^{1/\beta} \sigma_0))}{\int_t^\infty f(\tau| \beta, k^{1/\beta} \sigma_0))\,d\tau}
$$
represents the risk of a threshold crossing per unit time, and is a decreasing function for the Mittag-Leffler distribution. 
```{r hazard, fig.height=5, fig.cap="Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\\beta = 0.8$, $\\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\\beta$.\\label{fig:hazard}"}
tail <- 0.7
scale <- 1
t_0 <- c(0, 1, 10)
from <- 0.1
to <- 100 * scale
xx <- exp(seq(from = log(from), to = log(to), length.out = 100))
par(mfrow = c(1,2))
plot(xx, rep(1, length(xx)), ylim = c(0.001,1), type = 'n', log = 'xy', 
     xlab = 't', ylab = 'p')
for (k in 1:length(t_0)) {
  cond <- pml(q = t_0[k], tail = tail, scale = scale, lower.tail = FALSE)
  yy <- dml(x = xx + t_0[k], tail = tail, scale = scale) / cond
  lines(xx,yy, col = k+1, lty = k)
}
legend("bottomleft", c("t_0 = 0", "t_0 = 1", "t_0 = 10"), col = 2:4, lty = 1:3)

@@

To test our inference method, we have simulated independent waiting time and magnitude pairs $(W_k, J_k)$. The distribution of $J_k$ is irrelevant for the inference on $\beta$ and $\sigma_0$ (it just plays a role for the rescaling of $\hat{\sigma}_0$, which gets done to receive an estimate for the scale parameter of the exceedances). Hence we choosed exponential distributed magnitudes in all considered examples. For the waitings times, we choosed a Stable, a Pareto and an Exponential distribution.   
In order to have exact analytical values available for $\beta$ and $\sigma_0$, a distribution for $W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known. In our first example, we choose $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}, then due to the stability property we have the *equality* of distribution $W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, for $b(n) = n^{1/\beta}$. Using the parametrisation of @SamorodnitskyTaqqu, a few lines of calculation (see e.g. the vignette on parametrisation in @MittagLeffleR) show that $D$ must have the stable distribution $S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is implemented in the R package `stabledist` by @stabledist. If we choose $W_k$ to be Pareto distributed or more precisecly 
\begin{align*}
P(W_k>t)= C t^{-\beta}
\end{align*}
with $C:=(1/\Gamma(1-\beta))^{1/\beta}$ and $b(n)=n^{1/\beta}$ we get 
\begin{align*}
\frac{W_1 + \ldots + W_n}{b(n)} \stackrel{d}{\Rightarrow} D
\end{align*}
with $E(\exp(-sD))=\exp(-s^{\beta})$.

@@

# Data example

>>>>

Since the exponential distribution is nested in the Mittag-Leffler family of distributions, an appropriate way to choose between a model with exponential and Mittag-Leffler inter-exceedance times seems to be a Likelihood ratio test. Although the two models are nested, the assymptotic distribution is not $\chi^2_1$-distributed, the classical Theorem of Wilk doesn't apply since under $H_0$ the parameter $\beta$ of the Mittag-Leffler distribution is equal to one and hence lies under $H_0$ on the boundary of the parameter space $(0,1]$. Nonetheless, one can perform a bootstrapped log-likelihood ratio test. In Figure \ref{Fig:LRT} one can see the power for the bootstrapped LRT for Mittag-Leffler distributions with different tail parameters. As expected the power gets smaller for tail parameters close to one, since the Mittag-Leffler distribution converges for $\beta \rightarrow 1$ to an exponential distribution. Consequently, for tail parameters close to one it is hard to differentiate a Mittag-Lefller distribution from an exponential.

@@

The Theorem in Section \ref{sec:scaling} implies that for a high 
threshold $\ell$ we may approximate the distribution of $T(\ell)$ with an 
${\rm ML}(\beta, b(1/p))$ distribution, 
where the function $b(c)$ varies regularly at $\infty$ with parameter $1/\beta$.
Building on the POT (Peaks Over Threshold) method, we propose the following estimation procedure for the distribution of inter-exceedance time $T(\ell)$:

@@

According to Figure \ref{fig:flares}, for a threshold $\ell$ at the $k$-th order 
statistic, the fitted threshold exceedance time distribution is 
$$
T(\ell) \sim {\rm ML}(\beta, k^{-1/\beta} \sigma_0), 
$$
where $\beta = 0.85$ and $\sigma_0 = 3.0 \times 10^7 {\rm sec}$. 
Unlike the exponential distribution, the Mittag-Leffler distribution is not memoryless, and the probability density of the time $t$ until the next threshold crossing will depend on the time $t_0$ elapsed since the last threshold crossing.
This density equals 
$$
p(t|\beta, \sigma_0, \ell, t_0) = \frac{f(t + t_0 | \beta, k^{-1/\beta} \sigma_0)}{\mathbf P[T_\ell > t_0]}
$$
where $f(\,\cdot\, | \beta, k^{-1/\beta} \sigma_0)$ is the probability density of ${\rm ML}(\beta, -k^{1/\beta} \sigma_0)$. 
The more time has passed without a threshold crossing, the more the probability distribution shifts towards larger values for the next crossing
(see Figure \ref{fig:hazard}, left panel). The hazard rate 
$$
h(t) = \frac{f(t| \beta, k^{-1/\beta} \sigma_0))}{\int_t^\infty f(\tau| \beta, k^{-1/\beta} \sigma_0))\,d\tau}
$$
represents the risk of a threshold crossing per unit time, and is a decreasing function for the Mittag-Leffler distribution. 
```{r hazard, fig.height=5, fig.cap="Left: Conditional distribution of time until next threshold crossing, depending on elapsed time $t_0$ since last crossing ($\\beta = 0.8$, $\\sigma_0 = 1$). Right: Hazard rate depending on tail parameter $\\beta$.\\label{fig:hazard}"}
tail <- 0.7
scale <- 1
t_0 <- c(0, 1, 10)
from <- 0.1
to <- 100 * scale
xx <- exp(seq(from = log(from), to = log(to), length.out = 100))
par(mfrow = c(1,2))
plot(xx, rep(1, length(xx)), ylim = c(0.001,1), type = 'n', log = 'xy', 
     xlab = 't', ylab = 'p')
for (k in 1:length(t_0)) {
  cond <- pml(q = t_0[k], tail = tail, scale = scale, lower.tail = FALSE)
  yy <- dml(x = xx + t_0[k], tail = tail, scale = scale) / cond
  lines(xx,yy, col = k+1, lty = k)
}
legend("bottomleft", c("t_0 = 0", "t_0 = 1", "t_0 = 10"), col = 2:4, lty = 1:3)

@@

To test our inference method, we have simulated independent waiting time and magnitude pairs $(W_k, J_k)$. The distribution of $J_k$ is irrelevant for the inference on $\beta$ and $\sigma_0$ (it just plays a role for the rescaling of $\hat{\sigma}_0$, which gets done to receive an estimate for the scale parameter of the exceedances). Hence we choosed exponential distributed magnitudes in all considered examples. For the waitings times, we choosed (i) a Stable, (ii) a Pareto and (iii) an Exponential distribution.   
In order to have exact analytical values available for $\beta$ and $\sigma_0$, a distribution for $W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known. In our first example, we choose $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}, then due to the stability property we have the *equality* of distribution $W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, for $b(n) = n^{1/\beta}$. Using the parametrisation of @SamorodnitskyTaqqu, a few lines of calculation (see e.g. the vignette on parametrisation in @MittagLeffleR) show that $D$ must have the stable distribution $S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is implemented in the R package `stabledist` by @stabledist. If we choose $W_k$ to be Pareto distributed or more precisecly 
\begin{align*}
P(W_k>t)= C t^{-\beta}
\end{align*}
with $C:=(1/\Gamma(1-\beta))^{1/\beta}$ and $b(n)=n^{1/\beta}$ we get 
\begin{align*}
\frac{W_1 + \ldots + W_n}{b(n)} \stackrel{d}{\Rightarrow} D
\end{align*}
with $E(\exp(-sD))=\exp(-s^{\beta})$.

@@

```{r Ex1figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex1figs} Stability Plots for m=100 simulation runs for Example 1 (stable distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Tail08_logMoms.pdf","SimuStudy_StabilityPlots_Tail08_mles.pdf","SimuStudy_StabilityPlots_Tail08_hill.pdf"))
```

```{r Ex2figs, echo=FALSE,fig.height=7,fig.cap="\\label{Fig:Ex2figs} Stability Plots for m=100 simulation runs for Example 2 (Pareto distributed waiting times, true tail parameter =0.8). The grey lines, the red line and the blue lines are the log Moment, the MLE and the hill estimator, repectively, for different $k$'s on the x-axis. The dark lines in each picture are the means of the concerning estimators for different $k$'s.", fig.show='hold', out.width = '32%'}
knitr::include_graphics(c("SimuStudy_StabilityPlots_Ex2_Tail08_logMoms.pdf","SimuStudy_StabilityPlots_Ex2_Tail08_mles.pdf","SimuStudy_StabilityPlots_Ex2_Tail08_hill.pdf"))
```
# Data example
